### 2019-09-18 ch1_introduction/content.tex
update

%%%% HT: I feel like you're only going to focus on preprocessing, when reading this section. Maybe add a second paragraph mentioning that separating technical and biological noise from actual biological processes in the data is not the only challenge, and that there is also a necessity of new computational tools to start generating more complex models, integrating the information from single cell high sequencing experiments in a much more refined way?
%%% RC: In general, these two paragraphs are not telling the reader a lot about where this thesis is going

@@

This work makes incremental contributions to the field in order to be able to address the aforementioned problems in a more comprehensive context. This chapter first introduces several key concepts in both cell biology and computer science, upon which the remainder of this work relies. Afterwards, the research objectives and main contributions of this work are outlined.
% TODO: explain that only the main concepts are introduced?

\subsection{The origins of life and the RNA world}
The discovery of the double helix shape of Deoxyribonucleic Acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution. By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism. With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body.
The magnitude of this discovery is reflected in our language and culture alike; with sayings such as "It's in your DNA.", or usage of its shape in countless illustrations or artworks (Figure \ref{fig:fsvm}).
%% HT: I'm not sure if the magnitude of the DNA discovery is best reflected in sayings and double helix monuments it sounds a bit light to me, compared to the fact that you can identify geneticel targets of a disease and Crispr cas it for instance, but that's a personal opinion, feel free to ignore it :)

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/fsvm3} % TODO: change image
	\caption{
		\textbf{A prominent display of the double helix shape at the VIB FSVM building.} % TODO: expand description
	}
	\label{fig:fsvm}
\end{figure}

Even so, a widely-accepted hypothesis states that life (or cells) did not originate from DNA, but instead was kicked off from its lesser-known cousin, Ribonucleic Acid (RNA). According to the RNA world hypothesis \cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}). Only later did cells develop the ability to use DNA and proteins to self-sustain in a process commonly referred to as the Central Dogma.
% TODO: use this reference? \cite{cornell_prebioticaminoacids_2019}

@@

\subsection{The Central Dogma}
The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existent living cells. In short, it states that DNA codes for RNA, which in turn codes for proteins.
In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure 3).

%% TODO: add figure

@@

A stretch of DNA that contains a genetic blueprint for a particular molecule is called a gene, and the collection of an organism's genes is called its genome. \textbf{Transcription} is the process of synthesising an RNA molecule from a gene, and the resulting molecule is called a transcript. RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand.
% HT: , which is able to exit the nucleus compartment.
% RC: I haven't introduced the concept of a nucleus yet
Due to its single-strandedness, RNA is less stable and will break down faster. Single-strandedness also allows some types of RNA (e.g. transfer RNA, ribosomal RNA) to form more complex three-dimensional structures by having certain regions bind to other regions of the strand. This work only considers messenger RNA (mRNA), which are transcribed from protein-coding genes, meaning that the mRNAs can result in the production of particular proteins. A protein-coding gene consists of alternating sections called introns and exons. Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
% TODO: mention that there are about 20'000 protein-coding genes in the human body.

@@

% TODO: do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.
Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task. In this regard, it makes sense not only to reason about cell types but also about the transition states between cell types and the dynamic processes involved therein.
% TODO: are dynamic processes defined? a dynamic process is just a biochemical reaction

\subsection{Cell dynamics and gene regulation}
If cells are dynamic entities and can gradually produce the molecules needed to acquire new functionality, what is the process by which this happens? The mechanism by which this happens is called gene regulation. Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate). Such proteins are called transcription factors (TFs), and the genes they regulate are called their targets. Typically, one TF will regulate the transcription rate of many targets. % TODO: download database and calculate some statistics.

@@

% TODO: explain regulation mechanisms such as transcription factor binding sites?
\subsection{Profiling single cells}
In order to understand a biological process, it is often quite helpful to be able to profile (i.e. observe) the biomolecular components involved therein. The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.

The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
% HT: I wouldn't write that if it was my thesis, but then again, maybe it's only my personal choice
Since it was developed by Coons et al. in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins \cite{coons_immunologicalpropertiesantibody_1941}. In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system. A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.

@@

% HT: Is the section above needed? I felt like this 1.1.5 section was quite hard to read compared to the light previous sections.

On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting \cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner. There are many types of "omics", but the most commonly used are the following. In genomics, all of an organism's genes are studied -- its whole genome. Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively. A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics. If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable. Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts. By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
% Demonstrate the masking effect of bulk analyses.

@@

% TODO: provide better connection to the subsections below

\subsection{Normalisation}
% TODO: need to write still

@@

There are many ways of classifying DR methods \cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible. Examples of commonly used projection-based DR methods in single-cell omics are PCA and MDS. 
Manifold learning methods are methods which reconstruct a higher-order structure in the original space (e.g. a graph or a grid), visualising the structure in a lower-dimensional space, and mapping the original samples to the lower-dimensional space. Manifold learning can be an iterative optimisation process using a predefined criterion. Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP. 
% section is not very interestingly written

@@

@@

\section{Outline} % include contributions?
% mention focus on datasets which contain developing cells.

This work aims to tackle multiple computational problems in extracting biological hypotheses from single-cell omics data. (TODO: rephrase that we focus on datasets that contain developing cells, and that the tools we develop make this assumption.)
We introduce a framework for evaluating multiple types of single-cell omics tools, more specifically dimensionality reduction, trajectory inference, and gene regulatory network inference (Chapter \ref{chap:dyngen}). Next, we perform a comprehensive evaluation of 45 TI methods and construct guidelines based on our observations (Chapter \ref{chap:dynbenchmark}). As part of the evaluation, we developed a framework to infer, visualise, and interpret trajectories, which we made available as a separate tool called dyno (Chapter \ref{chap:dyno}).
We also introduce two new computational tools of our own. SCORPIUS is a TI method specialised in inferring linear trajectories (Chapter \ref{chap:scorpius}). bred is a true single-cell NI method which can infer gene regulatory networks for individual cells (Chapter \ref{chap:bred}).

@@

@@

We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends. This includes help speed up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes). Several of these package receive millions of downloads per year (devtools, remotes, rlang).

@@

		\cranpkg{diffusionMap} & aut & CRAN & 21361 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\

>>>>

\comment{
HT: I feel like you're only going to focus on preprocessing, when reading this section. Maybe add a second paragraph mentioning that separating technical and biological noise from actual biological processes in the data is not the only challenge, and that there is also a necessity of new computational tools to start generating more complex models, integrating the information from single cell high sequencing experiments in a much more refined way? \\
RC: Literally but succinctly state what the exact contributions of the thesis are, this is still very vague.
}

@@

This work makes incremental contributions to the field in order to be able to address the aforementioned problems in a more comprehensive context. This chapter first introduces several key concepts in both cell biology and computer science, upon which the remainder of this work relies. Afterwards, the research objectives and main contributions of this work are outlined.

\subsection{The origins of life and the RNA world}
The discovery of the double helix shape of Deoxyribonucleic Acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution. By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism. With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body.
The magnitude of this discovery is reflected in our language and culture alike; with sayings such as "It's in your DNA.", or usage of its shape in countless illustrations or artworks (Figure \ref{fig:fsvm}).

\comment{
RC: Use crispr-cas to assert the impact of the discovery of DNA, not the FSVM spiral 
}

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/fsvm3} % TODO: change image
	\caption{
		\textbf{A prominent display of the double helix shape at the VIB FSVM building.}
	}
	\label{fig:fsvm}
\end{figure}

Even so, a widely-accepted hypothesis states that life (or cells) did not originate from DNA, but instead was kicked off from its lesser-known cousin, Ribonucleic Acid (RNA). According to the RNA world hypothesis \cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}). Only later did cells develop the ability to use DNA and proteins to self-sustain in a process commonly referred to as the Central Dogma.

\comment{use this reference? \cite{cornell_prebioticaminoacids_2019}}

@@

\subsection{The Central Dogma}
\comment{HT: feels like an enumeration, should make it more into a story}

The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existent living cells. In short, it states that DNA codes for RNA, which in turn codes for proteins.
In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure \ref{fig:centraldogma}).

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/central_dogma} % TODO: change image
	\caption{
		\textbf{Central Dogma.} % TODO: expand the description
	}
	\label{fig:centraldogma}
\end{figure}

@@

A stretch of DNA that contains a genetic blueprint for a particular molecule is called a gene, and the collection of an organism's genes is called its genome. \textbf{Transcription} is the process of synthesising an RNA molecule from a gene, and the resulting molecule is called a transcript. RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand.
% HT: Add: which is able to exit the nucleus compartment.
% RC: I haven't introduced the concept of a nucleus yet
Due to its single-strandedness, RNA is less stable and will break down faster. Single-strandedness also allows some types of RNA (e.g. transfer RNA, ribosomal RNA) to form more complex three-dimensional structures by having certain regions bind to other regions of the strand. This work only considers messenger RNA (mRNA), which are transcribed from protein-coding genes, meaning that the mRNAs can result in the production of particular proteins. A protein-coding gene consists of alternating sections called introns and exons. Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
% TODO: mention that there are about 20'000 protein-coding genes in the human body.

@@

\comment{do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.}
Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task. In this regard, it makes sense not only to reason about cell types but also about the transition states between cell types and the dynamic processes involved therein.

\comment{are dynamic processes defined? a dynamic process is just a biochemical reaction}

\subsection{Cell dynamics and gene regulation}
If cells are dynamic entities and can gradually produce the molecules needed to acquire new functionality, what is the process by which this happens? The mechanism by which this happens is called gene regulation. Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate). Such proteins are called transcription factors (TFs), and the genes they regulate are called their targets. Typically, one TF will regulate the transcription rate of many targets.

\comment{TODO: download database and calculate some statistics?}

@@

\comment{TODO: explain regulation mechanisms such as transcription factor binding sites?}
\subsection{Profiling single cells}
\comment{This sections goes greatly into detail about IHC and cytometry. Can the merging of the 'single-cell' and 'omics' be explained without going so much into detail?}
In order to understand a biological process, it is often quite helpful to be able to profile (i.e. observe) the biomolecular components involved therein. The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.

The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
Since it was developed by Coons et al. in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins \cite{coons_immunologicalpropertiesantibody_1941}. In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system. A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.

\comment{Rephrase the microscope sentence}

@@

On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting \cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner. There are many types of "omics", but the most commonly used are the following. In genomics, all of an organism's genes are studied -- its whole genome. Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively. A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics. If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable. Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts. By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
\comment{Demonstrate the masking effect of bulk analyses.}

@@

\comment{Make a better connection to the subsections below}

\subsection{Normalisation}
\comment{TODO}

@@

There are many ways of classifying DR methods \cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible. Examples of commonly used projection-based DR methods in single-cell omics are PCA and MDS. 
Manifold learning methods are methods which reconstruct a higher-order structure in the original space (e.g. a graph or a grid), visualising the structure in a lower-dimensional space, and mapping the original samples to the lower-dimensional space. Manifold learning can be an iterative optimisation process using a predefined criterion. Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.

\comment{This section is not very interestingly written}

@@

\comment{Could still expand this section with pieces from the EJI paper, though it needs to be adapted strongly.}

@@

\section{Outline} % include contributions?

\comment{Get a better focus for the overall contributions of the thesis. For example, focus on datasets that contain developing cells, and that the tools we develop make this assumption.}

This work aims to tackle multiple computational problems in extracting biological hypotheses from single-cell omics data. 
We introduce a framework for evaluating multiple types of single-cell omics tools, more specifically dimensionality reduction, trajectory inference, and gene regulatory network inference (Chapter \ref{chap:dyngen}). Next, we perform a comprehensive evaluation of 45 TI methods and construct guidelines based on our observations (Chapter \ref{chap:dynbenchmark}). As part of the evaluation, we developed a framework to infer, visualise, and interpret trajectories, which we made available as a separate tool called dyno (Chapter \ref{chap:dyno}).
We also introduce two new computational tools of our own. SCORPIUS is a TI method specialised in inferring linear trajectories (Chapter \ref{chap:scorpius}). bred is a true single-cell NI method which can infer gene regulatory networks for individual cells (Chapter \ref{chap:bred}).

@@

\comment{
	Submitted:\\
	RNA atlas,\\
	TRADE-seq: Trajectory-based differential expression analysis,\\
	SCENIC protocol
}

@@

We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends. This includes help speed up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes). Several of these package receive millions of downloads per year (devtools, remotes, rlang).

@@

		\cranpkg{diffusionMap} & aut & CRAN & 21'361 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
